---
title: "Wytham Masting Study — Descriptive Analysis"
subtitle: "Non-genomic analyses"
author: "Wytham Masting Project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    number_sections: true
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  warning    = FALSE,
  message    = FALSE,
  fig.width  = 8,
  fig.height = 5,
  dpi        = 300,
  out.width  = "100%"
)

# Set working directory to R1_revision root so all relative paths work
# If knitting from within the Rmd/ subfolder, adjust accordingly:
# knitr::opts_knit$set(root.dir = "../")
```

# Overview

This document reproduces all **descriptive analyses** for the manuscript
*"Environmental and genomic associations of inter-annual and inter-tree
variation in acorn production in pedunculate oak"*.

Figures produced here correspond to:

| Figure | Description |
|--------|-------------|
| Fig. 1a | Annual acorn counts — visual method, per tree |
| Fig. 1b–f | Annual reproductive material — litter traps |
| Fig. 2 | % contribution per tree to total annual production |
| Fig. 3 | Correlation: visual counts vs. litter trap mature acorns |
| Table S1 | Descriptive statistics for all response variables |
| Table S2 | Population-level variability, synchrony, autocorrelation |

---

# Packages and paths

```{r packages}
library(tidyverse)     # data manipulation and ggplot2
library(readxl)        # read Excel files
library(ggthemes)      # extra ggplot2 themes
library(viridis)       # colour palettes
library(geomtextpath)  # smoothed labelled lines (Fig. 3)
library(plotrix)       # std.error()
library(car)           # Anova() for type-II tests
library(scales)        # scale formatting
library(patchwork)     # multi-panel figures

# ── Paths (relative to R1_revision/) ──────────────────────────────────────
DATA_RAW <- file.path("..", "data", "raw")
FIG_OUT  <- file.path("..", "outputs", "figures")
TAB_OUT  <- file.path("..", "outputs", "tables")

dir.create(FIG_OUT, recursive = TRUE, showWarnings = FALSE)
dir.create(TAB_OUT, recursive = TRUE, showWarnings = FALSE)

source("proj_theme.R")
select <- dplyr::select
filter <- dplyr::filter
```

---

# Data loading

## Visual counts (mature acorn counts, survey method)

Visual counts were made annually by counting acorns on each tree via
binoculars. Data span 2020–2023. Tree "ET" (extra tree) is excluded from
main analyses as it was not part of the core 38-tree panel.

```{r load-viz}
viz_raw <- read_excel(file.path(DATA_RAW, "visual_counts_2020_2023.xlsx"),
                      sheet = "Sheet1")

# Parse year from Date; remove ET
viz <- viz_raw %>%
  mutate(year = as.factor(format(as.Date(Date), "%Y")),
         Tree = as.factor(Tree)) %>%
  filter(Tree != "ET")

cat("Visual count data: ", nrow(viz), "rows |",
    "Years:", paste(levels(viz$year), collapse = ", "), "\n")
glimpse(viz)
```

## Litter trap data

Litter traps (0.25 m² per trap, six traps per tree) were deployed from
2021 onwards. **No litter trap data exist for 2020** — traps were not
in place during that season. This explains the absence of 2020 data in
Fig. 1b–f.

```{r load-traps}
trap <- read_excel(file.path(DATA_RAW, "litter_traps_2021_2023.xlsx"),
                   sheet = "Sheet1") %>%
  mutate(Year = as.factor(Year),
         Tree = as.factor(Tree))

cat("Litter trap data:", nrow(trap), "rows |",
    "Years:", paste(levels(trap$Year), collapse = ", "), "\n")
glimpse(trap)
```

## Full dataset (tree-level explanatory + response variables)

This is the master dataset used for statistical modelling (Section 03).
Loaded here for descriptive statistics.

```{r load-full}
full_df <- read_excel(file.path(DATA_RAW, "wytham_full_dataset.xlsx"),
                      sheet = "Data") %>%
  filter(TREE_ID != "ET") %>%
  mutate(
    DBH           = as.numeric(DBH),
    VIZ_COUNT     = as.numeric(VIZ_COUNT),   # stored as character in Excel
    PRODUCER      = as.factor(PRODUCER),
    SOIL_CLASS    = as.factor(SOIL_CLASS),
    across(c(CLAY, SILT, SAND, Mg_MGL, K_MGL, P_MGL, PH), as.numeric)
  ) %>%
  # Remove nutrient index columns (retained as mg/L concentrations only)
  dplyr::select(-any_of(c("K_INDEX", "P_INDEX", "Mg_INDEX")))

cat("Full dataset:", nrow(full_df), "rows |",
    nlevels(as.factor(full_df$TREE_ID)), "trees\n")
```

---

# Descriptive statistics table for response variables

Response variables defined:

| Variable | Definition | Units |
|----------|-----------|-------|
| VIZ_COUNT | Mature acorn count, visual method | count |
| MATURE_ACORNS | Mature acorns in litter traps (per 0.25 m²) | count |
| IMMAT_ACORNS | Immature acorns in litter traps | count |
| ABORTED_ACORNS | Enlarged cups (aborted acorns) in litter traps | count |
| FLOWERS | Catkin/flower counts in litter traps | count |
| CROP_PERCENT | % of total annual acorn crop | % |

```{r desc-stats-table}
resp_vars <- c("VIZ_COUNT", "MATURE_ACORNS", "IMMAT_ACORNS",
               "ABORTED_ACORNS", "FLOWERS", "CROP_PERCENT")

desc_stats <- full_df %>%
  # Use one row per tree (the full_df may be tree-averaged already)
  distinct(TREE_ID, .keep_all = TRUE) %>%
  select(all_of(resp_vars)) %>%
  pivot_longer(everything(), names_to = "Variable") %>%
  group_by(Variable) %>%
  summarise(
    N      = sum(!is.na(value)),
    Mean   = mean(value, na.rm = TRUE),
    SD     = sd(value, na.rm = TRUE),
    Median = median(value, na.rm = TRUE),
    Min    = min(value, na.rm = TRUE),
    Max    = max(value, na.rm = TRUE),
    CV_pct = SD / Mean * 100,
    .groups = "drop"
  ) %>%
  mutate(across(where(is.numeric), ~ round(.x, 2))) %>%
  arrange(match(Variable, resp_vars))

# Rename for readability
desc_stats_display <- desc_stats %>%
  rename(
    `Variable`   = Variable,
    `N`          = N,
    `Mean`       = Mean,
    `SD`         = SD,
    `Median`     = Median,
    `Min`        = Min,
    `Max`        = Max,
    `CV (%)`     = CV_pct
  )

knitr::kable(desc_stats_display,
             caption = "**Table S1.** Descriptive statistics for response
             variables measured in 38 pedunculate oak trees at Wytham Woods.
             CV = coefficient of variation (SD/Mean × 100). Note that
             MATURE_ACORNS, IMMAT_ACORNS, ABORTED_ACORNS and FLOWERS are
             averaged across litter trap collections within year. The high CVs
             justify the use of a Gamma (log link) GLM rather than Gaussian
             models (Gamma is appropriate for skewed, strictly positive
             continuous data).",
             align = "lrrrrrrr")

# Save to outputs
write.csv(desc_stats_display,
          file.path(TAB_OUT, "Table_R1_descriptive_statistics_response_vars.csv"),
          row.names = FALSE)
```

---

# Figure 1: Annual variation in acorn production

## Fig. 1a — Visual counts by year

Trees 27 (consistently poor producer) and 36 (consistently super-producer)
are marked with distinct point shapes and labelled.

```{r fig1a, fig.height=5, fig.width=6, fig.cap="Fig. 1a. Acorn counts per tree per year at Wytham Woods (visual method). Grey boxplots show the population distribution; individual trees shown as jittered points. Trees 36 (super-producer) and 27 (poor producer) are highlighted."}
# Tag focal trees
viz_tagged <- viz %>%
  mutate(focal = case_when(
    Tree == "36" ~ "Tree 36 (super producer)",
    Tree == "27" ~ "Tree 27 (poor producer)",
    TRUE         ~ "Other trees"
  ),
  focal = factor(focal, levels = c("Tree 36 (super producer)",
                                   "Tree 27 (poor producer)",
                                   "Other trees")))

fig1a <- ggplot(viz_tagged, aes(x = year, y = Counts)) +
  geom_boxplot(outlier.shape = NA, fill = "grey90", colour = "grey40",
               width = 0.5) +
  geom_jitter(data = filter(viz_tagged, focal == "Other trees"),
              colour = "grey60", size = 1.5, alpha = 0.7,
              width = 0.15, height = 0) +
  geom_point(data = filter(viz_tagged, focal != "Other trees"),
             aes(shape = focal, colour = focal),
             size = 2, stroke = 1.2) +
  scale_shape_manual(values = c(17, 15)) +
  scale_colour_manual(values = c("firebrick", "steelblue4")) +
  labs(x = "", y = "Acorn count (visual)",
       shape = "", colour = "") +
  guides(shape  = guide_legend(nrow = 2),
         colour = guide_legend(nrow = 2)) +
  proj_theme +
  theme(legend.position  = "top",
        legend.margin    = margin(t = -20),
        legend.spacing.y = unit(0, "pt"),
        legend.key.height = unit(10, "pt"),
        panel.grid = element_blank())

fig1a

ggsave(file.path(FIG_OUT, "Fig1a_visual_counts_by_year.pdf"),
       fig1a, width = 90, height = 100, units = "mm")
ggsave(file.path(FIG_OUT, "Fig1a_visual_counts_by_year.png"),
       fig1a, width = 60, height = 60, units = "mm", dpi = 300)
```

## Fig. 1b–f — Litter trap components by year

Litter traps were not deployed until 2021; the 2020 masting event is
captured only by the visual count method (Fig. 1a). This limitation is
explicitly stated in the figure caption.

```{r fig1b-f-panels, fig.height=8, fig.width=10, fig.cap="Fig. 1b–f. Annual counts of reproductive and associated material from litter traps (0.25 m² per trap) under 38 pedunculate oak trees at Wytham Woods, 2021–2023. Note: litter traps were not deployed in 2020; that year's acorn production is captured by visual counts (Fig. 1a)."}

# Reshape to long format for faceted plotting
trap_long <- trap %>%
  pivot_longer(cols = c(Acorns, Immature_Acorns, Enlarged_cup, Flower, Gall),
               names_to = "Category", values_to = "Count") %>%
  mutate(Category = factor(Category,
    levels = c("Acorns", "Immature_Acorns", "Enlarged_cup", "Flower", "Gall"),
    labels = c("Mature acorns", "Immature acorns",
               "Enlarged cups\n(aborted)", "Flowers",
               "Galls")))

fig1b_f <- ggplot(trap_long, aes(x = Year, y = Count)) +
  geom_boxplot(outlier.shape = NA, fill = "grey90", colour = "grey40",
               width = 0.5) +
  geom_jitter(colour = "black", size = 1.2, alpha = 0.7,
              width = 0.15, height = 0) +
  facet_wrap(~ Category, scales = "free_y", ncol = 3, axes = "all") +
  labs(x = "", y = "Count per trap (0.25 m²)") +
  proj_theme +
  theme(panel.grid = element_blank())

fig1b_f

ggsave(file.path(FIG_OUT, "Fig1b-f_litter_trap_components.pdf"),
       fig1b_f, width = 170, height = 120, units = "mm")
ggsave(file.path(FIG_OUT, "Fig1b-f_litter_trap_components.png"),
       fig1b_f, width = 120, height = 120, units = "mm", dpi = 300)
```

---

# Figure 2: Total acorn production per tree with annual deviations

Trees are ordered left to right by cumulative total count (highest producer
Tree 36 on left). Top panel: stacked raw counts by year. Lower panels:
deviation of each tree from the yearly population mean, one panel per year
(2021 excluded — no acorns matured).

```{r fig2-counts-deviation, fig.height=10, fig.width=12, fig.cap="Fig. 2. Total number of acorns produced by individual pedunculate oak trees at Wytham Woods over 2020, 2022, 2023 from visual counts (top); no acorns matured in 2021. Lower panels show deviation of each tree from the yearly mean acorn count. Trees are ranked from highest (Tree 36) to lowest (Tree 27) total production."}

# NEJM palette for years (2021 excluded from stacked bars — zero production)
nejm_4 <- c("2020" = "#BC3C29",
             "2022" = "#0072B5",
             "2023" = "#20854E")

# Order trees by cumulative total count
tree_order <- viz %>%
  group_by(Tree) %>%
  summarise(cum_count = sum(Counts, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(cum_count)) %>%
  pull(Tree)

viz_ordered <- viz %>%
  mutate(Tree = factor(Tree, levels = tree_order),
         year = as.character(year))

# ── Top panel: stacked raw counts (exclude 2021 — all zero) ─────────────
viz_nonzero <- viz_ordered %>% filter(year != "2021")

p_top <- ggplot(viz_nonzero, aes(x = Tree, y = Counts, fill = year)) +
  geom_col(width = 0.85, colour = NA) +
  scale_fill_manual(values = nejm_4, name = NULL) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.03))) +
  labs(x = NULL, y = "Acorn counts") +
  guides(fill = guide_legend(nrow = 1, keywidth = 0.8, keyheight = 0.5)) +
  proj_theme +
  theme(axis.text.x  = element_blank(),
        axis.ticks.x = element_blank(),
        legend.position = c(0.98, 0.98),
        legend.justification = c(1, 1),
        legend.background = element_rect(fill = "white", colour = NA))

# ── Deviation panels: one per year ───────────────────────────────────────
deviation_panels <- function(yr) {
  d <- viz_ordered %>%
    filter(year == yr) %>%
    group_by(year) %>%
    mutate(deviation = Counts - mean(Counts, na.rm = TRUE)) %>%
    ungroup()

  ggplot(d, aes(x = Tree, y = deviation)) +
    geom_col(fill = "grey70", colour = "grey40", width = 0.8, linewidth = 0.2) +
    geom_hline(yintercept = 0, colour = "black", linewidth = 0.4) +
    annotate("text", x = Inf, y = Inf, label = yr,
             hjust = 1.1, vjust = 1.4, size = 3.5, fontface = "bold") +
    scale_y_continuous(limits = c(-150, 350)) +
    labs(x = NULL, y = NULL) +
    proj_theme +
    theme(axis.text.x = element_text(size = 7, colour = "black"))
}

p_2020 <- deviation_panels("2020") + theme(axis.text.x = element_blank(),
                                            axis.ticks.x = element_blank())
p_2022 <- deviation_panels("2022") +
  labs(y = "Deviation from yearly mean acorn count") +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
p_2023 <- deviation_panels("2023") +
  labs(x = "Tree")

fig2 <- p_top / p_2020 / p_2022 / p_2023 +
  plot_layout(heights = c(1.4, 1, 1, 1))

fig2

ggsave(file.path(FIG_OUT, "Fig2_acorn_counts_and_deviations.pdf"),
       fig2, width = 150, height = 120, units = "mm")
ggsave(file.path(FIG_OUT, "Fig2_acorn_counts_and_deviations.png"),
       fig2, width = 150, height = 120, units = "mm", dpi = 300)

# Headline numbers
top7_trees <- tree_order[1:7]
top7_pct <- viz %>%
  mutate(year_total = ave(Counts, year, FUN = function(x) sum(x, na.rm=TRUE)),
         pct = Counts / year_total * 100) %>%
  filter(Tree %in% top7_trees) %>%
  group_by(year) %>%
  summarise(top7_pct = sum(pct, na.rm = TRUE))
cat("Top 7 trees % of annual total by year:\n")
print(top7_pct)
```

---

# Figure 3: Correlation between visual counts and litter trap acorns

Visual counts and litter trap mature acorn counts are compared to validate
the two methods. Trees 3, 6, 10, 32, 34, 36 are excluded as they lack
complete data across methods.

```{r fig3-correlation, fig.height=5, fig.width=6, fig.cap="Fig. 3. Correlation between visual acorn counts and litter trap mature acorn counts per 0.25 m² for individual trees, 2021–2023 (2020 excluded as litter traps not deployed). Lines fitted by ordinary least squares per year."}
EXCLUDE_TREES <- c("3", "6", "10", "32", "34", "36")

viz_cor <- viz %>%
  filter(!Tree %in% EXCLUDE_TREES, year != "2020") %>%
  rename(Year = year)

trap_cor <- trap %>%
  filter(!Tree %in% EXCLUDE_TREES) %>%
  rename(Acorns_trap = Acorns)

combined <- inner_join(viz_cor, trap_cor, by = c("Tree", "Year"))

# Kendall and Pearson correlations reported in-text
kt <- cor.test(combined$Counts, combined$Acorns_trap, method = "kendall")
pr <- cor.test(log(combined$Counts + 1), log(combined$Acorns_trap + 1),
               method = "pearson")

cat("Kendall tau =", round(kt$estimate, 3), ", p =", round(kt$p.value, 4), "\n")
cat("Pearson r (log+1) =", round(pr$estimate, 3), ", p =", round(pr$p.value, 4), "\n")

nejm_2 <- c("2022" = "#0072B5", "2023" = "#20854E")

fig3 <- ggplot(combined, aes(x = Counts, y = Acorns_trap,
                              shape = Year, linetype = Year, colour = Year)) +
  geom_point(size = 2.5) +
  geom_smooth(aes(group = Year), method = "lm", formula = y ~ x,
              se = FALSE, linewidth = 0.8) +
  scale_shape_manual(values = c("2022" = 17, "2023" = 15)) +
  scale_linetype_manual(values = c("2022" = "solid",
                                   "2023" = "dashed")) +
  scale_colour_manual(values = nejm_2) +
  guides(shape    = guide_legend(title = NULL),
         linetype = guide_legend(title = NULL),
         colour   = guide_legend(title = NULL)) +
  labs(x = "Visual acorn count",
       y = "Mature acorns\nper trap (0.25 m²)") +
  proj_theme +
  theme(legend.position  = "top",
        legend.margin    = margin(t = 0, b = 0),
        legend.spacing.y = unit(0, "pt"),
        legend.key.height = unit(10, "pt"),
        legend.key.width  = unit(12, "pt"),
        panel.grid = element_blank())

fig3

ggsave(file.path(FIG_OUT, "Fig3_visual_vs_trap_correlation.pdf"),
       fig3, width = 90, height = 90, units = "mm")
ggsave(file.path(FIG_OUT, "Fig3_visual_vs_trap_correlation.png"),
       fig3, width = 65, height = 60, units = "mm", dpi = 300)
```

---

# Population-level reproductive variability, autocorrelation and synchrony

With only 4 years of data (3 non-zero years for litter traps), all estimates
of autocorrelation and synchrony carry considerable uncertainty. Results
should be interpreted as preliminary descriptors rather than formal
statistical tests.

```{r variability-synchrony}
# ── Individual-level CV ──────────────────────────────────────────────────
# Use NA for trees with mean_count = 0 (super-poor producers in all years)
indiv_cv <- viz %>%
  group_by(Tree) %>%
  summarise(
    mean_count = mean(Counts, na.rm = TRUE),
    sd_count   = sd(Counts, na.rm = TRUE),
    CV_pct     = ifelse(mean_count > 0, sd_count / mean_count * 100, NA_real_),
    n_years    = n(),
    .groups    = "drop"
  ) %>%
  arrange(desc(CV_pct))

cat("Individual-level CV (visual counts across all years):\n")
cat("  Mean CV:", round(mean(indiv_cv$CV_pct, na.rm = TRUE), 1), "%\n")
cat("  Median CV:", round(median(indiv_cv$CV_pct, na.rm = TRUE), 1), "%\n")
cat("  Range:", round(min(indiv_cv$CV_pct, na.rm = TRUE), 1), "–",
    round(max(indiv_cv$CV_pct, na.rm = TRUE), 1), "%\n\n")

# ── Population CV per year ───────────────────────────────────────────────
# 2021 was a near-complete masting failure; if all counts = 0, CV is undefined
pop_cv <- viz %>%
  group_by(year) %>%
  summarise(
    mean_count = mean(Counts, na.rm = TRUE),
    sd_count   = sd(Counts, na.rm = TRUE),
    CV_pct     = ifelse(mean_count > 0, sd_count / mean_count * 100, NA_real_),
    .groups    = "drop"
  )

cat("Population CV by year (inter-individual variation):\n")
print(pop_cv)
cat("  Note: NA indicates a near-zero year where CV is undefined (mean ≈ 0).\n")

# ── Population-level variability (inter-annual CV of population mean) ────
pop_annual <- viz %>%
  group_by(year) %>%
  summarise(mean_count = mean(Counts), .groups = "drop")

cv_interannual <- sd(pop_annual$mean_count) / mean(pop_annual$mean_count) * 100
cat("\nInter-annual CV of population mean:", round(cv_interannual, 1), "%\n")

# ── Synchrony: Pearson correlation among all tree pairs ─────────────────
# Pivot to wide format (tree × year)
viz_wide <- viz %>%
  pivot_wider(id_cols = Tree, names_from = year, values_from = Counts)

# Compute pairwise Pearson correlations across years (requires ≥3 years)
sync_pairs <- combn(nrow(viz_wide), 2, function(idx) {
  x <- as.numeric(viz_wide[idx[1], -1])
  y <- as.numeric(viz_wide[idx[2], -1])
  # Use complete pairs only
  ok <- complete.cases(x, y)
  if (sum(ok) < 3) return(NA_real_)
  cor(x[ok], y[ok], method = "pearson")
})

mean_sync <- mean(sync_pairs, na.rm = TRUE)
cat("\nPopulation synchrony (mean pairwise Pearson r across all tree pairs):\n")
cat("  Mean r =", round(mean_sync, 3), "\n")
cat("  SD r =", round(sd(sync_pairs, na.rm = TRUE), 3), "\n")
cat("  Note: based on only", sum(!is.na(sync_pairs)), "pairs with ≥3 shared years.\n")
cat("  ⚠ With n = 4 years this estimate is unreliable; report with caution.\n\n")

# ── Lag-1 autocorrelation (per tree, then averaged) ──────────────────────
lag1_ac <- viz %>%
  arrange(Tree, year) %>%
  group_by(Tree) %>%
  summarise(
    ac1 = {
      x <- Counts
      # Require at least 3 time points and non-constant series
      if (length(x) >= 3 && sd(x, na.rm = TRUE) > 0 &&
          sd(x[-length(x)], na.rm = TRUE) > 0)
        suppressWarnings(cor(x[-length(x)], x[-1], method = "pearson"))
      else NA_real_
    },
    .groups = "drop"
  )

cat("Lag-1 autocorrelation of individual trees:\n")
cat("  Mean:", round(mean(lag1_ac$ac1, na.rm = TRUE), 3), "\n")
cat("  SD:", round(sd(lag1_ac$ac1, na.rm = TRUE), 3), "\n")
cat("  ⚠ With n = 4 time points this is highly unreliable; treat descriptively.\n")
cat("  ⚠ The strongly negative mean value is dominated by the 2020 (high) → 2021\n")
cat("    (near-zero) transition. With only 3 lag-1 pairs per tree, this single\n")
cat("    transition drives the estimate. Do not over-interpret the magnitude.\n")
```

```{r variability-table}
# Summary table for manuscript (Table R2)
table_r2 <- tibble(
  Metric = c(
    "Mean individual CV (%)",
    "Population CV — 2020 (%)",
    "Population CV — 2021 (%) [mast failure]",
    "Population CV — 2022 (%)",
    "Population CV — 2023 (%)",
    "Inter-annual CV of population mean (%)",
    "Mean pairwise synchrony (Pearson r)",
    "Mean lag-1 autocorrelation (Pearson r)"
  ),
  Value = c(
    round(mean(indiv_cv$CV_pct, na.rm = TRUE), 1),
    round(filter(pop_cv, year == "2020")$CV_pct, 1),
    ifelse(is.na(filter(pop_cv, year == "2021")$CV_pct), "NA (mean ≈ 0)",
           as.character(round(filter(pop_cv, year == "2021")$CV_pct, 1))),
    round(filter(pop_cv, year == "2022")$CV_pct, 1),
    round(filter(pop_cv, year == "2023")$CV_pct, 1),
    round(cv_interannual, 1),
    round(mean_sync, 3),
    round(mean(lag1_ac$ac1, na.rm = TRUE), 3)
  ),
  Notes = c(
    "Mean across 38 trees; all years",
    "Inter-individual variation within year",
    "Inter-individual variation within year",
    "Inter-individual variation within year",
    "Inter-individual variation within year",
    "Year-to-year variation in population mean",
    "Mean of all pairwise r; n=4 yrs, use with caution",
    "Mean across trees; n=4 time points, very limited"
  )
)

knitr::kable(table_r2,
             caption = "**Table S2.** Reproductive variability and synchrony
             statistics for 38 pedunculate oak trees at Wytham Woods.
             CV = coefficient of variation. Synchrony estimated as mean
             pairwise Pearson correlation coefficient (Loreau & de Mazancourt
             2008 approach). Lag-1 autocorrelation reported per tree then
             averaged. All estimates based on n = 4 years of data and should
             be interpreted with caution.",
             align = "lrr")

write.csv(table_r2,
          file.path(TAB_OUT, "Table_R2_variability_synchrony.csv"),
          row.names = FALSE)
```

---

# Super/medium/low producer classification excluding 2020

Producer categories are re-derived on 2021–2023 data to assess whether
the 2020 mast year drives the classification.

```{r superproducer-no2020}
# Load full dataset PRODUCER labels (original classification on all 4 years)
full_prod <- read_excel(file.path(DATA_RAW, "wytham_full_dataset.xlsx"),
                        sheet = "Data") %>%
  filter(TREE_ID != "ET") %>%
  select(TREE_ID, PRODUCER) %>%
  distinct() %>%
  mutate(Tree = as.character(TREE_ID))

# Visual counts excluding 2020
viz_no2020 <- viz %>%
  filter(year != "2020") %>%
  group_by(Tree) %>%
  summarise(
    mean_count_no2020 = mean(Counts, na.rm = TRUE),
    sum_count_no2020  = sum(Counts,  na.rm = TRUE),
    .groups = "drop"
  )

# Re-classify using the same quantile thresholds applied to the no-2020 means
q33 <- quantile(viz_no2020$mean_count_no2020, 1/3, na.rm = TRUE)
q67 <- quantile(viz_no2020$mean_count_no2020, 2/3, na.rm = TRUE)

viz_no2020 <- viz_no2020 %>%
  mutate(
    Producer_no2020 = case_when(
      mean_count_no2020 >= q67 ~ "Super",
      mean_count_no2020 >= q33 ~ "Medium",
      TRUE                     ~ "Low"
    ),
    Producer_no2020 = factor(Producer_no2020,
                             levels = c("Super", "Medium", "Low"))
  )

# Merge with original classification
prod_compare <- full_prod %>%
  left_join(viz_no2020, by = "Tree") %>%
  mutate(PRODUCER = as.character(PRODUCER),
         Same_class = (PRODUCER == Producer_no2020))

# Summary: how many trees changed class?
cat("Producer classification with vs. without 2020:\n")
cat("  Total trees:", nrow(prod_compare), "\n")
cat("  Same class:", sum(prod_compare$Same_class, na.rm = TRUE), "\n")
cat("  Changed class:", sum(!prod_compare$Same_class, na.rm = TRUE), "\n\n")

# Cross-tabulation
cross_tab <- table(Original = prod_compare$PRODUCER,
                   No2020   = prod_compare$Producer_no2020)
cat("Cross-tabulation (rows = original, cols = without 2020):\n")
print(cross_tab)

# Trees that changed
changed <- prod_compare %>%
  filter(!Same_class) %>%
  select(Tree, PRODUCER, Producer_no2020, mean_count_no2020)
if (nrow(changed) > 0) {
  cat("\nTrees that changed classification:\n")
  print(changed)
} else {
  cat("\n✓ All trees retain the same producer classification when 2020 is excluded.\n")
}

knitr::kable(prod_compare %>%
               select(Tree, Original = PRODUCER, No2020 = Producer_no2020,
                      Mean_no2020 = mean_count_no2020, Same_class) %>%
               arrange(Original, No2020),
             caption = "**Producer classification with and without 2020.**
             Original: classification from wytham_full_dataset.xlsx (all years).
             No2020: reclassified on 2021–2023 means using tertile thresholds.
             Same_class = TRUE if the producer category did not change.",
             digits = 1)

write.csv(prod_compare %>%
            select(Tree, TREE_ID, Original_PRODUCER = PRODUCER,
                   Producer_no2020, mean_count_no2020, sum_count_no2020,
                   Same_class),
          file.path(TAB_OUT, "Superproducer_without_2020.csv"),
          row.names = FALSE)
```

---

# Session information

```{r session-info}
sessionInfo()
```
