---
title: "Wytham Masting Study — Microclimate Analysis"
subtitle: "Non-genomic analyses"
author: "Wytham Masting Project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    number_sections: true
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  warning   = FALSE,
  message   = FALSE,
  fig.width = 9,
  fig.height= 6,
  dpi       = 300,
  out.width = "100%"
)
```

# Overview

This document covers all **microclimate analyses** for the manuscript,
corresponding to Supplementary Figures 1–3 and Supplementary Table 9.

| Output | Description |
|--------|-------------|
| SI Fig. 1 | Heatmap of daily mean temperature (2020–2023) |
| SI Fig. 2 | Heatmap of daily precipitation (2020–2023) |
| SI Fig. 3 | Individual tree deviations from mean microclimate |
| SI Table 9 | ANOVAs: microclimate ~ producer class × phenological period × year |

---

# Packages and paths

```{r packages}
library(tidyverse)
library(lubridate)
library(viridis)
library(car)       # Anova() type II
library(readxl)

# car loads MASS which masks dplyr::select — restore it
select <- dplyr::select
filter <- dplyr::filter

DATA_RAW <- file.path("..", "data", "raw")
FIG_OUT  <- file.path("..", "outputs", "figures")
TAB_OUT  <- file.path("..", "outputs", "tables")

dir.create(FIG_OUT, recursive = TRUE, showWarnings = FALSE)
dir.create(TAB_OUT, recursive = TRUE, showWarnings = FALSE)

source("proj_theme.R")
```

---

# Data loading

## Tree microclimate logger data

Logger data were cleaned in `Logger_cleaning.R` (archived in `obsolete/`)
and saved as an RDS file containing daily mean, maximum, and minimum
temperatures per tree. Loggers were deployed in October 2020.

```{r load-mc}
mc <- readRDS(file.path(DATA_RAW, "tree_microclimate_daily_2020_2023.rds"))

mc <- mc %>%
  mutate(
    Date = as.Date(Date),
    Tree = as.factor(Tree),
    year = year(Date),
    month = month(Date),
    day   = day(Date)
  ) %>%
  filter(Tree != "ET")

cat("Microclimate data:", nrow(mc), "rows\n")
cat("Date range:", format(min(mc$Date)), "to", format(max(mc$Date)), "\n")
cat("Trees:", nlevels(mc$Tree), "\n")
```

## Precipitation data

Daily precipitation from the UK Environmental Change Network (ECN)
Wytham station (data courtesy of Stefanie M. Schäfer and Denise Palle).

```{r load-precip}
precip <- read_csv(file.path(DATA_RAW, "wytham_daily_precipitation_2020_2023.csv"),
                   col_names = c("Date", "Rainfall_mm"),
                   show_col_types = FALSE) %>%
  mutate(
    Date       = dmy(Date),
    Rainfall_mm = suppressWarnings(as.numeric(Rainfall_mm)),
    year   = year(Date),
    month  = month(Date),
    day    = day(Date)
  ) %>%
  filter(!is.na(Date))

cat("Precipitation data:", nrow(precip), "rows |",
    format(min(precip$Date)), "to", format(max(precip$Date)), "\n")
```

## Producer class classification

Used for SI Table 9 grouping.

```{r load-producers}
full_df <- read_excel(file.path(DATA_RAW, "wytham_full_dataset.xlsx"),
                      sheet = "Data") %>%
  filter(TREE_ID != "ET") %>%
  select(TREE_ID, PRODUCER) %>%
  distinct() %>%
  mutate(Tree     = as.factor(TREE_ID),
         PRODUCER = as.factor(PRODUCER))

mc <- mc %>% left_join(full_df, by = "Tree")
```

---

# SI Figure 1: Daily mean temperature heatmap

The heatmap spans the full calendar years 2020–2023. Logger data begins October 2020 (logger deployment date); cells for January–September 2020 are left blank (NA) in the per-tree data.

```{r si-fig1-temp-heatmap, fig.height=7, fig.width=10, fig.cap="SI Fig. 1. Heatmap of daily mean temperature (°C) averaged across all tree-mounted loggers, 2020–2023. Grey cells indicate dates prior to logger deployment (loggers deployed October 2020) or dates with no logger data. The global colour scale spans -5.5°C to +28°C."}

# Population-mean daily temperature across trees
mc_pop_daily <- mc %>%
  group_by(Date, year, month, day) %>%
  summarise(
    Temp_mean = mean(Temperature, na.rm = TRUE),
    n_trees   = sum(!is.na(Temperature)),
    .groups   = "drop"
  )

# Expand grid to include all calendar dates 2020–2023
all_dates <- tibble(Date = seq(as.Date("2020-01-01"),
                               as.Date("2023-12-31"), by = "day")) %>%
  mutate(year = year(Date), month = month(Date), day = day(Date))

mc_full_grid <- all_dates %>%
  left_join(mc_pop_daily, by = c("Date", "year", "month", "day"))

# Global temperature range for consistent scale
t_min <- min(mc_pop_daily$Temp_mean, na.rm = TRUE)
t_max <- max(mc_pop_daily$Temp_mean, na.rm = TRUE)

# One heatmap per year
plot_year_heatmap <- function(df, yr, legend = FALSE) {
  p <- df %>%
    filter(year == yr) %>%
    ggplot(aes(x = day, y = month, fill = Temp_mean)) +
    geom_tile() +
    scale_fill_viridis(option = "H", direction = 1,
                       limits = c(t_min, t_max),
                       name = "°C", na.value = "grey90") +
    scale_x_continuous(limits = c(0.5, 31.5),
                       breaks = seq(1, 31, by = 5)) +
    scale_y_continuous(limits = c(0.5, 12.5),
                       breaks = 1:12,
                       labels = month.abb) +
    labs(title = yr, x = "Day", y = "") +
    proj_theme +
    theme(panel.grid = element_blank(),
          axis.text.x = element_text(size = 10))

  if (!legend) p <- p + guides(fill = "none")
  p
}

library(patchwork)

p_2020 <- plot_year_heatmap(mc_full_grid, 2020)
p_2021 <- plot_year_heatmap(mc_full_grid, 2021)
p_2022 <- plot_year_heatmap(mc_full_grid, 2022)
p_2023 <- plot_year_heatmap(mc_full_grid, 2023, legend = TRUE)

si_fig1 <- (p_2020 | p_2021) / (p_2022 | p_2023)
si_fig1

ggsave(file.path(FIG_OUT, "SI_Fig1_temperature_heatmap_2020_2023.pdf"),
       si_fig1, width = 170, height = 140, units = "mm")
ggsave(file.path(FIG_OUT, "SI_Fig1_temperature_heatmap_2020_2023.png"),
       si_fig1, width = 170, height = 140, units = "mm", dpi = 300)
```

---

# SI Figure 2: Daily precipitation heatmap

```{r si-fig2-precip-heatmap, fig.height=7, fig.width=10, fig.cap="SI Fig. 2. Heatmap of daily total precipitation (mm) at Wytham Woods, 2020–2023, from the UK Environmental Change Network (ECN) station."}

prec_min <- 0
prec_max <- max(precip$Rainfall_mm, na.rm = TRUE)

plot_precip_heatmap <- function(df, yr, legend = FALSE) {
  p <- df %>%
    filter(year == yr) %>%
    ggplot(aes(x = day, y = month, fill = Rainfall_mm)) +
    geom_tile() +
    scale_fill_viridis(option = "D", direction = -1,
                       limits = c(prec_min, prec_max),
                       name = "mm", na.value = "grey90") +
    scale_x_continuous(limits = c(0.5, 31.5),
                       breaks = seq(1, 31, by = 5)) +
    scale_y_continuous(limits = c(0.5, 12.5),
                       breaks = 1:12,
                       labels = month.abb) +
    labs(title = yr, x = "Day", y = "") +
    proj_theme +
    theme(panel.grid = element_blank(),
          axis.text.x = element_text(size = 10))

  if (!legend) p <- p + guides(fill = "none")
  p
}

p2_2020 <- plot_precip_heatmap(precip, 2020)
p2_2021 <- plot_precip_heatmap(precip, 2021)
p2_2022 <- plot_precip_heatmap(precip, 2022)
p2_2023 <- plot_precip_heatmap(precip, 2023, legend = TRUE)

si_fig2 <- (p2_2020 | p2_2021) / (p2_2022 | p2_2023)

si_fig2

ggsave(file.path(FIG_OUT, "SI_Fig2_precipitation_heatmap_2020_2023.pdf"),
       si_fig2, width = 170, height = 140, units = "mm")
ggsave(file.path(FIG_OUT, "SI_Fig2_precipitation_heatmap_2020_2023.png"),
       si_fig2, width = 170, height = 140, units = "mm", dpi = 300)
```

---

# SI Figure 3: Individual tree microclimate deviations

Differences between each individual tree's daily mean/max/min temperature
and the population mean, ordered by acorn production rank (high to low).

```{r tree-order}
# Order trees by total visual count (needs visual count data)
viz <- read_excel(file.path(DATA_RAW, "visual_counts_2020_2023.xlsx"),
                  sheet = "Sheet1") %>%
  filter(Tree != "ET") %>%
  mutate(Tree = as.character(Tree))

tree_rank <- viz %>%
  group_by(Tree) %>%
  summarise(total_acorns = sum(Counts, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(total_acorns)) %>%
  mutate(rank = row_number())

tree_order_vec <- tree_rank$Tree
```

```{r si-fig3-deviations, fig.height=9, fig.width=12, fig.cap="SI Fig. 3. Difference of maximum, minimum and mean temperature for 38 individual pedunculate oak trees at Wytham Woods, Oxford from means of the whole sample. The x-axis is ordered by tree from high to low numbers of total acorns produced in 2020–2023."}

# Per-tree mean of daily mean, max, min temperature across all logger days
tree_temp_means <- mc %>%
  filter(!is.na(Tree)) %>%
  group_by(Tree) %>%
  summarise(
    Mean = mean(Temperature,         na.rm = TRUE),
    Max  = mean(Maximum.Temperature, na.rm = TRUE),
    Min  = mean(Minimum.Temperature, na.rm = TRUE),
    .groups = "drop"
  )

# Population mean for each metric
pop_means <- tree_temp_means %>%
  summarise(across(c(Mean, Max, Min), \(x) mean(x, na.rm = TRUE)))

# Deviation per tree per metric
mc_dev <- tree_temp_means %>%
  mutate(
    Mean = Mean - pop_means$Mean,
    Max  = Max  - pop_means$Max,
    Min  = Min  - pop_means$Min,
    Tree = factor(as.character(Tree), levels = tree_order_vec)
  ) %>%
  pivot_longer(c(Max, Mean, Min),
               names_to  = "Metric",
               values_to = "deviation") %>%
  mutate(Metric = factor(Metric, levels = c("Max", "Mean", "Min"),
                         labels = c("Maximum temperature",
                                    "Mean temperature",
                                    "Minimum temperature")))

si_fig3 <- ggplot(mc_dev, aes(x = Tree, y = deviation, fill = deviation)) +
  geom_col() +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey30") +
  scale_fill_gradient2(low = "steelblue", mid = "white", high = "firebrick",
                       midpoint = 0, name = "Deviation (°C)") +
  facet_wrap(~ Metric, ncol = 1) +
  labs(
    x = "Tree (ordered by total acorn production, high → low)",
    y = "Deviation from whole-sample mean (°C)"
  ) +
  proj_theme +
  theme(
    axis.text.x     = element_text(size = 7, angle = 45, hjust = 1),
    panel.grid      = element_blank(),
    legend.position = "right"
  )

si_fig3

ggsave(file.path(FIG_OUT, "SI_Fig3_individual_tree_temp_deviations.pdf"),
       si_fig3, width = 170, height = 110, units = "mm")
ggsave(file.path(FIG_OUT, "SI_Fig3_individual_tree_temp_deviations.png"),
       si_fig3, width = 170, height = 110, units = "mm", dpi = 300)
```

---

# SI Table 9: ANOVAs for microclimate × producer class

These linear models test whether trees classified as high, average, or poor
acorn producers differ in their microclimate, whether microclimate differed
across years, and whether there are producer class × year interactions.

```{r si-table9-anova}
# Seasonal averages per tree per year
mc_seasonal <- mc %>%
  filter(!is.na(PRODUCER)) %>%
  mutate(
    season = case_when(
      month %in% 4:5  ~ "Budburst",
      month %in% 6:8  ~ "AcornDev",
      month %in% 10:12 ~ "PostLeafDrop",
      TRUE             ~ NA_character_
    )
  ) %>%
  filter(!is.na(season)) %>%
  group_by(Tree, PRODUCER, year, season) %>%
  summarise(
    mean_temp = mean(Temperature, na.rm = TRUE),
    max_temp  = mean(Maximum.Temperature, na.rm = TRUE),
    min_temp  = mean(Minimum.Temperature, na.rm = TRUE),
    .groups   = "drop"
  ) %>%
  mutate(
    year    = as.factor(year),
    season  = factor(season,
                     levels = c("Budburst", "AcornDev", "PostLeafDrop"),
                     labels = c("Budburst (Apr–May)",
                                "Acorn dev. (Jun–Aug)",
                                "Post-leaf-drop (Oct–Dec)"))
  )

# ANOVA function (type II)
run_mc_anova <- function(resp_var, label) {
  formula <- as.formula(
    paste(resp_var, "~ PRODUCER * year + season + PRODUCER:season")
  )
  m <- lm(formula, data = mc_seasonal)
  av <- Anova(m, type = "II")
  av_df <- as.data.frame(av) %>%
    rownames_to_column("Term") %>%
    mutate(Response = label) %>%
    rename(`Sum Sq` = `Sum Sq`, `Df` = Df, `F value` = `F value`, `Pr(>F)` = `Pr(>F)`)
  av_df
}

anova_results <- bind_rows(
  run_mc_anova("mean_temp", "Mean temperature"),
  run_mc_anova("max_temp",  "Maximum temperature"),
  run_mc_anova("min_temp",  "Minimum temperature")
)

anova_display <- anova_results %>%
  filter(Term != "Residuals") %>%   # Residuals row has no F/p — exclude from display
  mutate(
    `Sum Sq`   = round(`Sum Sq`, 3),
    `F value`  = round(`F value`, 3),
    `Pr(>F)`   = ifelse(`Pr(>F)` < 0.001, "<0.001", as.character(round(`Pr(>F)`, 4)))
  ) %>%
  dplyr::select(Response, Term, Df, `Sum Sq`, `F value`, `Pr(>F)`)

knitr::kable(anova_display,
             caption = "**SI Table 9.** Results of linear model ANOVAs
             (Type II SS) testing the effects of acorn producer class
             (High/Average/Poor), phenological season, year, and their
             interactions on daily mean, maximum, and minimum temperatures
             of 38 pedunculate oak trees at Wytham Woods.",
             align = "llrrrr")

write.csv(anova_display,
          file.path(TAB_OUT, "SI_Table9_microclimate_ANOVA.csv"),
          row.names = FALSE)
```

---

# Session information

```{r session-info}
sessionInfo()
```
